### 2024-09-25 paper/paper.qmd
Add files via upload

---
title: "An Analysis of Subway Delays in Toronto: Insights from 2023 TTC Data"
subtitle: "Diving deep into subway delays"
author: 
  - Charlie Zhang
thanks: "Code and data are available at: https://github.com/Zqyyk11/Term-Paper-1."
date: today
date-format: long
abstract: "This report presents an investigation into subway delays in Toronto using data from the Toronto Transit Commission (TTC) for the year 2023. The analysis examines patterns of delays across various subway stations, lines, and days of the week, with the goal of identifying critical factors that contribute to service disruptions. Key findings show that Monday and Thursday experienced the highest number of delays, with the Bloor-Danforth (BD) line being the most affected. These insights provide a data-driven foundation for improving operational efficiency and minimizing service interruptions in the future."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false
library(ggplot2)
library(here)
library(knitr)
library(opendatatoronto)
library(tidyverse)
library(dplyr)

cleaned_data = read_csv(here::here('data/analysis_data/cleaned_subway_delay_data.csv'))

```

# Introduction

Public transit plays a vital role in the daily lives of millions of people in urban centers around the world. In Toronto, the subway system, operated by the Toronto Transit Commission (TTC), serves as a backbone for many commuters. However, delays in subway services can cause significant inconvenience and disrupt the city’s flow of traffic. In 2023, Toronto’s subway system faced numerous delays, which are the focus of this report.

This analysis uses a comprehensive dataset of subway delays from the TTC to investigate the nature and extent of service disruptions. Specifically, the report aims to answer the following questions:

-   Which days of the week are most prone to delays?
-   What are the most affected subway lines and stations?
-   What times of day experience the most delays?

By answering these questions, the report seeks to provide actionable insights for improving the reliability of Toronto’s subway system, reducing the frequency and duration of delays, and enhancing the overall commuting experience.

The findings in this report are based on data-driven analyses and visualizations, which highlight delay trends throughout the year. This investigation serves as a foundation for further research into the causes of delays and offers guidance for policy makers and transit authorities in addressing key operational challenges.

# Data {#sec-data}

The data utilized in this report is sourced from Open Data Toronto and was imported using the opendatatoronto library [@gelfand2022]. The specific dataset used for analyzing TTC subway delays in Toronto is the "TTC Subway & SRT Train Service Delay Data" dataset [@toronto2024. All data analysis was conducted using R [@r2023], with the assistance of several key packages, including tidyverse [@wickham2019], dplyr [@wickham2023dplyr],here [@mueller2020] janitor [@firke2023], ggplot2 [@wickham2016], and knitr [@knitr].

To transform the raw data into a clean, analysis-ready format, several preprocessing steps were carried out using R. First, unnecessary columns that did not contribute to the analysis, such as metadata or irrelevant identifiers, were removed to streamline the dataset. Using the dplyr package, specific columns were selected based on relevance to subway delays, such as date, time, station, line, and delay duration. The data was further cleaned by handling any missing or inconsistent values with the janitor package to ensure data integrity. Any outliers or erroneous entries were also identified and either corrected or removed as appropriate. These steps allowed for the creation of a clean dataset, which provided a solid foundation for the subsequent analysis. The entire process was conducted through efficient R code, leveraging the functionality of packages like dplyr for filtering and selection, and janitor for data cleaning.

```{r,echo=FALSE}
library(ggplot2)

# Assuming your cleaned data is in 'subway_data' and it has a 'Day' column
day_delay_summary <- aggregate(cleaned_data$`Min Delay`, by=list(cleaned_data$Day), FUN=sum)

# Plotting the total delays by day of the week
ggplot(day_delay_summary, aes(x=Group.1, y=x)) +
  geom_bar(stat="identity", fill="skyblue") +
  labs(title="Total Delays by Day of the Week (in minutes)",
       x="Day of the Week", y="Total Delay (Minutes)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The bar chart shows the total minutes of delays across different days of the week, with Mondays experiencing the highest delay times at 10,494 minutes, followed closely by Thursday and Friday. These days likely face more operational challenges due to higher ridership at the start and end of the workweek. In contrast, weekends see fewer delays, possibly because of lower ridership and reduced service frequency, leading to less strain on the subway system. \\ \\

```{r,echo=FALSE}
# Aggregating the data by subway line
line_delay_summary <- aggregate(cleaned_data$`Min Delay`, by=list(cleaned_data$Line), FUN=sum)

# Plotting the total delays by subway line
ggplot(line_delay_summary, aes(x=Group.1, y=x)) +
  geom_bar(stat="identity", fill="salmon") +
  labs(title="Total Delays by Subway Line (in minutes)",
       x="Subway Line", y="Total Delay (Minutes)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The bar chart shows the total delay minutes per subway line, with the Yonge-University (YU) line experiencing the most delays, followed by the Bloor-Danforth (BD) line. The high delay times on the YU Line are likely due to its status as one of the busiest lines, serving downtown Toronto and major commuter hubs. The BD Line also faces substantial delays, reflecting its heavy operational load. In contrast, other lines such as the Sheppard (SRT) experience significantly fewer delays, likely due to lower ridership or reduced service frequency. \\ \\

```{r,echo=FALSE}
# Aggregating and selecting the top 10 stations with the most delays
station_delay_summary <- aggregate(cleaned_data$`Min Delay`, by=list(cleaned_data$Station), FUN=sum)
top_10_stations <- station_delay_summary[order(-station_delay_summary$x),][1:10,]

# Plotting total delays by top 10 stations
ggplot(top_10_stations, aes(x=Group.1, y=x)) +
  geom_bar(stat="identity", fill="lightgreen") +
  labs(title="Total Delays by Top 10 Stations (in minutes)",
       x="Station", y="Total Delay (Minutes)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The bar chart highlights the top 10 stations with the highest delay minutes, with Kennedy BD Station leading, followed by Eglinton and Kipling stations. Kennedy, as a major terminal on the Bloor-Danforth line, likely faces operational challenges contributing to its delays. Eglinton Station, undergoing significant construction due to transit expansions, also experiences high delays. Key transfer hubs like Kipling and Bloor, which handle large passenger volumes, further contribute to increased delays at these stations.

# Results

The analysis of subway delays in Toronto for 2023 reveals significant patterns across days of the week, subway lines, and specific stations. Mondays experienced the most delays, followed closely by Thursday and Friday, suggesting that the beginning and end of the workweek are the most challenging periods for subway operations. This pattern may be linked to increased ridership and operational pressures during these days. In terms of subway lines, the Yonge-University (YU) line recorded the highest delay minutes, likely due to its extensive coverage of key commuter areas, including downtown Toronto, where traffic and operational complexities are higher. The Bloor-Danforth (BD) line also saw substantial delays, reflecting its heavy usage and operational load. Meanwhile, other lines, such as the Sheppard (SRT), experienced fewer delays, which can be attributed to lower ridership and less frequent service. At the station level, Kennedy BD Station led in delay minutes, followed by Eglinton and Kipling stations, all of which are either major transfer hubs or subject to ongoing construction, leading to operational disruptions. The results highlight the need for focused operational improvements at key lines and stations, particularly during peak periods at the start and end of the workweek. \\

```{r,echo=FALSE}
# Install required packages if necessary
# install.packages("ggplot2")
library(ggplot2)
library(dplyr)
cleaned_subway_data <- cleaned_data

# Assuming cleaned_subway_data is already loaded, we'll remove all rows with zero delay
cleaned_subway_data <- cleaned_subway_data %>% 
  filter(`Min Delay` > 0)

# Create a new 'Category' column combining 'Line' and 'Day of the Week'
cleaned_subway_data <- cleaned_subway_data %>%
  mutate(Category = paste(Line, Day))

# Aggregate the data by 'Category'
combined_summary <- cleaned_subway_data %>%
  group_by(Category) %>%
  summarize(Delay = sum(`Min Delay`))

# Plot the combined data without zero delays
ggplot(combined_summary, aes(x = Category, y = Delay, fill = Category)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  labs(title = "Combined Subway Delay Analysis by Line and Day of the Week (2023)",
       x = "Line and Day of the Week", y = "Total Delay (Minutes)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  coord_flip()  # Flip for better readability if needed

```

# Discussion

The analysis of subway delays in Toronto for 2023 reveals several important insights into the performance of the city's subway system. Delays are more pronounced during the workweek, particularly on Mondays and Thursdays, suggesting that the operational strain is highest at the start and end of the week, likely due to increased commuter traffic. The Yonge-University (YU) and Bloor-Danforth (BD) lines experience the most delays, reflecting their roles as major transit arteries with high passenger volumes and frequent service demands. Stations like Kennedy and Eglinton, which are key transfer points and hubs of construction activity, are more prone to delays, pointing to infrastructure and operational challenges that exacerbate service interruptions.

The removal of zero-delay entries further emphasizes that certain stations and lines consistently face challenges, while others see minimal delays. This highlights potential areas for targeted improvements, particularly on high-traffic lines and during peak commuting days. To reduce delays, strategic investments in infrastructure and more efficient scheduling could mitigate the operational pressures observed on busy lines and at critical stations. In addition, proactive measures, such as real-time monitoring and passenger load management, could further improve service reliability. Overall, the data points to a need for a more flexible and robust transit system capable of handling fluctuations in demand while minimizing disruptions.

\newpage

# References

### 2024-09-25 scripts/00-simulate_data.R

@@ -1,19 +1,27 @@
 #### Preamble ####
-# Purpose: Simulates... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
-# License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Purpose: Simulates subway delay data.
+# Author: Charlie Zhang
+# Date: Sep 23th, 2024
+# Contact: zqycharlie.zhang@mail.utoronto.ca
+# License: Open Government License - Toronto
+# Pre-requisites: None
+# Any other information needed? None
+
 
 
 #### Workspace setup ####
 library(tidyverse)
-# [...UPDATE THIS...]
+
 
 #### Simulate data ####
-# [...ADD CODE HERE...]
+set.seed(123)
+n <- 100
+simulated_data <- data.frame(
+  date = seq(as.Date('2020-01-01'), by = "day", length.out = n),
+  transportation_usage = rpois(n, lambda = 200),
+  location = sample(c("Downtown", "Midtown", "Suburb"), n, replace = TRUE)
+)
 
-
+# Print sample data
+print(head(simulated_data))
 

### 2024-09-25 scripts/01-download_data.R

@@ -1,26 +1,39 @@
 #### Preamble ####
-# Purpose: Downloads and saves the data from [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
-# License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Purpose: Download the raw data from opendatatoronto.
+# Author: Charlie Zhang
+# Date: Sep 23th, 2024
+# Contact: zqycharlie.zhang@mail.utoronto.ca
+# License: Open Government License - Toronto
+# Pre-requisites: None
+# Any other information needed? None
+
 
 
 #### Workspace setup ####
 library(opendatatoronto)
-library(tidyverse)
-# [...UPDATE THIS...]
+library(dplyr)
 
 #### Download data ####
-# [...ADD CODE HERE TO DOWNLOAD...]
+# get package
+package <- show_package("996cfe8d-fb35-40ce-b569-698d51fc683b")
+package
 
+# List the resources available in this package
+resources <- list_package_resources(package)
 
+# View the names and formats of the resources
+resource_id <- resources %>%
+  filter(name == "ttc-subway-delay-data-2023", format == "XLSX") %>%
+  pull(id)
+
+# Download the Excel file and save it to the raw_data folder
+subway_data <- get_resource(resource_id)
 
 #### Save data ####
-# [...UPDATE THIS...]
 # change the_raw_data to whatever name you assigned when you downloaded it.
-write_csv(the_raw_data, "inputs/data/raw_data.csv") 
+write.csv(subway_data, "data/raw_data/ttc_subway_delay_data_2023.csv", row.names = FALSE)
+
+
+
 
          

### 2024-09-25 scripts/02-data_cleaning.R

@@ -1,44 +1,29 @@
 #### Preamble ####
-# Purpose: Cleans the raw plane data recorded by two observers..... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 6 April 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
-# License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Purpose: Cleans the data and removes any NA variables/useless columns.
+# Author: Charlie Zhang
+# Date: Sep 23th, 2024
+# Contact: zqycharlie.zhang@mail.utoronto.ca
+# License: Open Government License - Toronto
+# Pre-requisites: None
+# Any other information needed? None
 
 #### Workspace setup ####
 library(tidyverse)
+library(dplyr)
 
 #### Clean data ####
-raw_data <- read_csv("inputs/data/plane_data.csv")
+# Select only the useful columns
+cleaned_data <- subway_data %>%
+  select(Date, Time, Day, Station, `Min Delay`,`Min Gap`, Line)
 
-cleaned_data <-
-  raw_data |>
-  janitor::clean_names() |>
-  select(wing_width_mm, wing_length_mm, flying_time_sec_first_timer) |>
-  filter(wing_width_mm != "caw") |>
-  mutate(
-    flying_time_sec_first_timer = if_else(flying_time_sec_first_timer == "1,35",
-                                   "1.35",
-                                   flying_time_sec_first_timer)
-  ) |>
-  mutate(wing_width_mm = if_else(wing_width_mm == "490",
-                                 "49",
-                                 wing_width_mm)) |>
-  mutate(wing_width_mm = if_else(wing_width_mm == "6",
-                                 "60",
-                                 wing_width_mm)) |>
-  mutate(
-    wing_width_mm = as.numeric(wing_width_mm),
-    wing_length_mm = as.numeric(wing_length_mm),
-    flying_time_sec_first_timer = as.numeric(flying_time_sec_first_timer)
-  ) |>
-  rename(flying_time = flying_time_sec_first_timer,
-         width = wing_width_mm,
-         length = wing_length_mm
-         ) |> 
-  tidyr::drop_na()
+# Remove rows with missing values in the selected columns
+cleaned_data <- cleaned_data %>%
[…]

### 2024-09-25 scripts/03-test_data.R

@@ -1,15 +1,34 @@
 #### Preamble ####
-# Purpose: Tests... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
-# License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Purpose: Test the cleaned data.
+# Author: Charlie Zhang
+# Date: Sep 23th, 2024
+# Contact: zqycharlie.zhang@mail.utoronto.ca
+# License: Open Government License - Toronto
+# Pre-requisites: None
+# Any other information needed? None
 
 
 #### Workspace setup ####
 library(tidyverse)
-# [...UPDATE THIS...]
 
 #### Test data ####
+# Check for the number of NA values in each column
+colSums(is.na(cleaned_data))
+
+# Remove rows with any missing values
+cleaned_data <- na.omit(cleaned_data)
+
+# Check if any values in 'Min Delay' are negative
+incorrect_delays <- cleaned_data %>%
+  filter(`Min Delay` < 0)
+
+# Remove rows where 'Min Delay' is negative
+cleaned_data <- cleaned_data %>%
+  filter(`Min Delay` >= 0)
+
+# Summary of the cleaned data
+summary(cleaned_data)
+
+# Check the first few rows of the cleaned data
+head(cleaned_data)
+

### 2024-09-26 scripts/00-simulate_data.R
first update

@@ -12,7 +12,7 @@
 #### Workspace setup ####
 library(tidyverse)
 
-
+daivsadsda
 #### Simulate data ####
 set.seed(123)
 n <- 100

### 2024-09-27 paper/paper.qmd
Second update with suggestions taken from peer review.

@@
 ---
-title: "An Analysis of Subway Delays in Toronto: Insights from 2023 TTC Data"
+title: "An Investigation of Subway Delays in Toronto: Insights from 2023 TTC Data"
 subtitle: "Diving deep into subway delays"
@@
   - Charlie Zhang
-thanks: "Code and data are available at: https://github.com/Zqyyk11/Term-Paper-1."
-date: today
+thanks: "Code and data are available at: https://github.com/Zqyyk11/Investigation-on-Toronto-Subway-Delay."
+date: "2024-09-23"
 date-format: long
 abstract: "This report presents an investigation into subway delays in Toronto using data from the Toronto Transit Commission (TTC) for the year 2023.
-The analysis examines patterns of delays across various subway stations, lines, and days of the week, with the goal of identifying critical factors that contribute to service disruptions.
-Key findings show that Monday and Thursday experienced the highest number of delays, with the Bloor-Danforth (BD) line being the most affected.
+The analysis examines patterns of delays across various subway stations, lines, and days of the week, with the goal of identifying critical factors that contribute to subway delays.
+Key findings show that Monday experienced the highest number of delays, with the Bloor-Danforth (BD) line being the most affected.
 These insights provide a data-driven foundation for improving operational efficiency and minimizing service interruptions in the future."
 format: pdf
+fontsize: 12pt
+toc: true
 number-sections: true
@@
 ---
-
 ```{r}
@@
 library(dplyr)
+library(lubridate)
 
@@
 ```
-
 # Introduction
@@
 The specific dataset used for analyzing TTC subway delays in Toronto is the "TTC Subway & SRT Train Service Delay Data" dataset [@toronto2024.
-All data analysis was conducted using R [@r2023], with the assistance of several key packages, including tidyverse [@wickham2019], dplyr [@wickham2023dplyr],here [@mueller2020] janitor [@firke2023], ggplot2 [@wickham2016], and knitr [@knitr].
+All data analysis was conducted using R [@r2023], with the assistance of several key packages, including tidyverse [@wickham2019], dplyr [@wickham2023dplyr],here [@mueller2020] janitor [@firke2023], ggplot2 [@wickham2016], knitr [@knitr], and lubridate [@grolemund2011].
 
 To transform the raw data into a clean, analysis-ready format, several preprocessing steps were carried out using R.
-First, unnecessary columns that did not contribute to the analysis, such as metadata or irrelevant identifiers, were removed to streamline the dataset.
+First, unnecessary columns that did not contribute to the analysis, were removed to clean the dataset.
 Using the dplyr package, specific columns were selected based on relevance to subway delays, such as date, time, station, line, and delay duration.
@@
 Any outliers or erroneous entries were also identified and either corrected or removed as appropriate.
-These steps allowed for the creation of a clean dataset, which provided a solid foundation for the subsequent analysis.
+These steps allowed for the creation of a clean dataset, which provided a solid foundation for the analysis below.
 The entire process was conducted through efficient R code, leveraging the functionality of packages like dplyr for filtering and selection, and janitor for data cleaning.
 
+## Overview of the data
+
+```{r,echo=FALSE}
+#| label: "Toronto Subway Delays 2023 Data"
+#| fig-cap: "Sample of the Toronto subway's delay timer"
+#| echo: false
+
+# PREVIEW OF THE DATA
+cleaned_data[, 1:7] %>% head() %>% knitr::kable()
+```
+
+## Subway delays by day of the week
 ```{r,echo=FALSE}
@@
 
-The bar chart shows the total minutes of delays across different days of the week, with Mondays experiencing the highest delay times at 10,494 minutes, followed closely by Thursday and Friday.
-These days likely face more operational challenges due to higher ridership at the start and end of the workweek.
-In contrast, weekends see fewer delays, possibly because of lower ridership and reduced service frequency, leading to less strain on the subway system.
-\\ \\
+The study of subway delays on days of the week uncovers a noticeable trend linked to how weekdays and weekends unfold in terms of commuter patterns and travel demand shifts within the city's transit system.
+Observations show that Mondays encounter the most substantial delays and also other weekdays like Thursdays and Fridays due to heightened rider activity during the typical workweek when individuals travel to their workplaces.
+The uptick in passenger traffic during peak hours exerts pressure on subway operations by causing trains and busier platforms along, with prolonged boarding processes that collectively contribute to amplified service disruptions.
+During peak hours when there is demand for subway services issues like mechanical problems or signal interruptions tend to cause bigger disruptions compared to weekends.
+A reason weekends experience fewer delays possibly because fewer people use the subway due to less frequent service and reduced ridership.
+With a number of commuters relying on the subway during weekends the system runs more smoothly with less strain resulting in better train operations.
+Moreover weekend schedules often incorporate planned maintenance activities that have impact since there are fewer passengers traveling.
+The difference, between weekdays and weekends shows how commuter traffic affects the reliability of subway services.
+Weekdays experience frequent and severe delays during rush hour demands compared to weekends when operations are lighter and services run more smoothly.
 
+## Subway delays by subway line
 ```{r,echo=FALSE}
@@
 
-The bar chart shows the total delay minutes per subway line, with the Yonge-University (YU) line experiencing the most delays, followed by the Bloor-Danforth (BD) line.
-The high delay times on the YU Line are likely due to its status as one of the busiest lines, serving downtown Toronto and major commuter hubs.
-The BD Line also faces substantial delays, reflecting its heavy operational load.
-In contrast, other lines such as the Sheppard (SRT) experience significantly fewer delays, likely due to lower ridership or reduced service frequency.
-\\ \\
+The Yonge University (Y U) and Bloor Danforth (B D) subway routes encounter delays mainly because they are the most heavily used and vital transportation links in Toronto city center and surrounding regions.
+These subway lines pass through populated neighborhoods and major transit junctions like downtown Toronto serving as crucial lifelines for the millions of people who rely on them daily for their commutes.
+Specifically the Y U Line provides access to locations such, as Union Station, financial centers and popular spots leading to congestion during busy periods.
+The BD Line runs through the city connecting neighborhoods from east to west.
+Is essential for many commuters getting to work or school and carrying out their daily routines efficiently.
+The large number of passengers using these lines presents difficulties like longer boarding times and crowded platforms along with frequent train delays.
+Furthermore the substantial ridership on these routes amplifies the effects of disruptions as any delays have a cascading effect throughout the system given the heavy dependence, on these crucial lines.
+On the hand the Sheppard line (SRT) has fewer delays compared to others mainly because it caters to smaller communities and operates with less frequent service thus easing the burden, on infrastructure and operations.
 
+## Subway delays by top 10 stations
 ```{r,echo=FALSE}
@@
 
-The bar chart highlights the top 10 stations with the highest delay minutes, with Kennedy BD Station leading, followed by Eglinton and Kipling stations.
-Kennedy, as a major terminal on the Bloor-Danforth line, likely faces operational challenges contributing to its delays.
-Eglinton Station, undergoing significant construction due to transit expansions, also experiences high delays.
-Key transfer hubs like Kipling and Bloor, which handle large passenger volumes, further contribute to increased delays at these stations.
+The top ten stations experiencing the delays are mainly major terminals and transfer points in Torontos subway system – with Kennedy BD Station leading the list due to its operational intricacies and heavy passenger traffic flow through connections with other lines, like the Scarborough RT and the Bloor Danforth line.
+Currently at Eglinton Station construction is underway for the Eglinton Crosstown LRT project causing disruptions that affect passenger flow and train operations resulting in increased delays.
+Kipling and Bloor stations are also facing delays as they serve as transfer points for various subway and bus routes, with many commuters switching lines or modes of transportation.
+The clustering of travelers at these locations results in crowded platforms and increases the likelihood of delays since even small interruptions can lead to a domino effect that impacts the reliability of the service provided there.This scenario occurs mainly in stations serving as central transit points due to their passenger flow and intricate operations along, with occasional infrastructure issues that collectively add up to prolonged delay times.
+
+## Subway delays Peak and off peak hours
+```{r,echo=FALSE}
+# Convert the 'Time' column from character to proper time format using lubridate's hms function
+cleaned_data$Time <- hms::as_hms(cleaned_data$Time)
+
+# Define peak hours (7:00 - 9:00 and 16:00 - 19:00) as well as off-peak hours
+cleaned_data <- cleaned_data %>%
+  mutate(
+    period = case_when(
+      (hour(Time) >= 7 & hour(Time) < 9) | (hour(Time) >= 16 & hour(Time) < 19) ~ "Peak",
+      TRUE ~ "Off-Peak"
+    )
+  )
+
+# Summarize the total delay minutes for Peak and Off-Peak
+delay_summary <- cleaned_data %>%
+  group_by(period) %>%
+  summarise(total_delay_minutes = sum(`Min Delay`, na.rm = TRUE))
+
+# Plotting the total delay for Peak vs. Off-Peak
+ggplot(delay_summary, aes(x = period, y = total_delay_minutes, fill = period)) +
+  geom_bar(stat = "identity") +
+  labs(title = "Total Subway Delay Minutes: Peak vs Off-Peak Hours",
+       x = "Time Period",
+       y = "Total Delay Minutes") +
+  theme_minimal()
+
+```
+
+The examination of subway disruptions in Toronto indicates a contrast between busy and quieter times of the day when it comes to delays.
+With less crowded periods encountering much longer total delay durations compared to peak hours despite the higher number of passengers and operational demands during those busier times when the transit system seems to focus on efficiency and manages to have fewer delays than, during the less busy times.
+Delays are more common during quieter hours due to frequent services and maintenance work being carried out then to avoid disruptions during prime commute times also fewer resources being available could lead to longer delays when problems occur.
+It seems that the system handles high demand situations effectively but could possibly do better at reducing interruptions during quieter times to maintain service all day long.
 
@@
 
-The analysis of subway delays in Toronto for 2023 reveals significant patterns across days of the week, subway lines, and specific stations.
-Mondays experienced the most delays, followed closely by Thursday and Friday, suggesting that the beginning and end of the workweek are the most challenging periods for subway operations.
-This pattern may be linked to increased ridership and operational pressures during these days.
-In terms of subway lines, the Yonge-University (YU) line recorded the highest delay minutes, likely due to its extensive coverage of key commuter areas, including downtown Toronto, where traffic and operational complexities are higher.
-The Bloor-Danforth (BD) line also saw substantial delays, reflecting its heavy usage and operational load.
-Meanwhile, other lines, such as the Sheppard (SRT), experienced fewer delays, which can be attributed to lower ridership and less frequent service.
-At the station level, Kennedy BD Station led in delay minutes, followed by Eglinton and Kipling stations, all of which are either major transfer hubs or subject to ongoing construction, leading to operational disruptions.
-The results highlight the need for focused operational improvements at key lines and stations, particularly during peak periods at the start and end of the workweek.
-\\
+The analysis of Toronto subway delays was conducted across four key aspects: delays by day of the week, subway lines, stations, and peak vs.
+off-peak hours.
+Each of these aspects offers unique insights into the operational challenges faced by the Toronto Transit Commission (TTC) in managing subway services.
 
-```{r,echo=FALSE}
-# Install required packages if necessary
-# install.packages("ggplot2")
-library(ggplot2)
-library(dplyr)
-cleaned_subway_data <- cleaned_data
+Delays by Day of the Week: The analysis revealed that Mondays experience the highest total delay minutes, likely due to the surge of passengers at the start of the workweek.
+Both Thursday and Friday also show elevated delays, potentially reflecting increased traffic as commuters prepare for the weekend.
+Weekends, on the other hand, see fewer delays, which can be attributed to lower ridership and reduced service frequency.
 
-# Assuming cleaned_subway_data is already loaded, we'll remove all rows with zero delay
-cleaned_subway_data <- cleaned_subway_data %>% 
-  filter(`Min Delay` > 0)
+Delays by Subway Line:The Yonge-University (YU) and Bloor-Danforth (BD) lines experienced the most delays.
+As two of the busiest lines serving major commuter hubs in downtown Toronto, they face greater operational challenges and are more susceptible to disruptions due to overcrowding and high service frequency.
+In contrast, smaller lines, like the Sheppard and Scarborough RT, showed fewer delays, likely due to lower ridership and fewer trains.
 
-# Create a new 'Category' column combining 'Line' and 'Day of the Week'
-cleaned_subway_data <- cleaned_subway_data %>%
-  mutate(Category = paste(Line, Day))
+Delays by Subway Station: Kennedy Station on the Bloor-Danforth (BD) line had the highest recorded delays, followed by Eglinton and Kipling stations.
+As key transfer and terminal points, these stations handle large volumes of passengers, which can cause bottlenecks, particularly during rush hours.
+Eglinton Station, which is undergoing heavy construction due to transit expansion projects, is another hotspot for delays.
 
-# Aggregate the data by 'Category'
-combined_summary <- cleaned_subway_data %>%
-  group_by(Category) %>%
-  summarize(Delay = sum(`Min Delay`))
+Peak vs.
+Off-Peak Hours:The final analysis compared peak and off-peak hours, revealing that off-peak periods experienced significantly higher total delay minutes.
+This might be due to reduced service frequency during off-peak hours, which makes each delay more impactful.
+Scheduled maintenance, often performed during off-peak hours, and fewer operational resources could also contribute to the higher delay times.
 
-# Plot the combined data without zero delays
-ggplot(combined_summary, aes(x = Category, y = Delay, fill = Category)) +
-  geom_bar(stat = "identity", show.legend = FALSE) +
-  labs(title = "Combined Subway Delay Analysis by Line and Day of the Week (2023)",
-       x = "Line and Day of the Week", y = "Total Delay (Minutes)") +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
-  coord_flip()  # Flip for better readability if needed
-
-```
+The overall analysis gives an overview of subway disruptions in Toronto for the year 2023.The frequency of delays is greatly impacted by the day of the week – particularly Mondays and the latter part of the workweek witness the highest delays.The Yonge-University and Bloor-Danforth lines face delays owing to their popularity and frequent train services.
+Specific stations like Kennedy and Eglinton serve as congestion points.
+Accumulate more delay minutes due to their vital positions in the transit system.
+During times when there are passengers traveling such as off hours or low traffic periods of the day there tends to be more delays in terms of total minutes being experienced due, to possibly having fewer trains running and maintenance work taking place.
+The thorough analysis offers an overview of subway disruptions in Toronto for the year 2023 with delays being significantly impacted by various factors and influences within the system.
+The analysis points out areas for operational enhancements to concentrate on such as improving off hours service and handling delays at busy stations like Kennedy and Eglinton to enhance the overall efficiency of the system better.
+Likewise tackling the issues on subway lines like YU and BD could lead to fewer disruptions especially during peak hours, with heavy passenger traffic.
 

### 2024-09-27 scripts/00-simulate_data.R

@@ -12,7 +12,6 @@
 #### Workspace setup ####
 library(tidyverse)
 
-daivsadsda
 #### Simulate data ####
 set.seed(123)
 n <- 100

### 2024-09-27 paper/paper.qmd
Third update improving grammar and simulation improvements

@@
 ```{r,echo=FALSE}
-library(ggplot2)
 
-# Assuming your cleaned data is in 'subway_data' and it has a 'Day' column
+#Create the summary
 day_delay_summary <- aggregate(cleaned_data$`Min Delay`, by=list(cleaned_data$Day), FUN=sum)
@@
 
+\newpage
+
 # Results
@@
 
-The analysis of subway delays in Toronto for 2023 reveals several important insights into the performance of the city's subway system.
-Delays are more pronounced during the workweek, particularly on Mondays and Thursdays, suggesting that the operational strain is highest at the start and end of the week, likely due to increased commuter traffic.
-The Yonge-University (YU) and Bloor-Danforth (BD) lines experience the most delays, reflecting their roles as major transit arteries with high passenger volumes and frequent service demands.
-Stations like Kennedy and Eglinton, which are key transfer points and hubs of construction activity, are more prone to delays, pointing to infrastructure and operational challenges that exacerbate service interruptions.
+The examination of delays, in the Toronto subway system reveals the challenges of managing a public transportation network efficiently and reliably.
+A notable discovery is the increase in delays during busy hours compared to peak periods – a surprising trend that goes against the assumption of more disruptions during high demand peak hours.
+The information indicates that when there are trains running and passengers must wait longer during off hours due to reduced service frequency which cause delays to have an impact.
+Moreover scheduled maintenance activities that are typically carried out during these quieter times may worsen the situation by causing delays to build up progressively over time.
+This situation underscores a chance for the TTC to reassess its off hour schedules and resource distribution in order to minimize delays during periods of passenger traffic.
 
-The removal of zero-delay entries further emphasizes that certain stations and lines consistently face challenges, while others see minimal delays.
-This highlights potential areas for targeted improvements, particularly on high-traffic lines and during peak commuting days.
-To reduce delays, strategic investments in infrastructure and more efficient scheduling could mitigate the operational pressures observed on busy lines and at critical stations.
-In addition, proactive measures, such as real-time monitoring and passenger load management, could further improve service reliability.
-Overall, the data points to a need for a more flexible and robust transit system capable of handling fluctuations in demand while minimizing disruptions.
+During hours the transportation system faces stress but appears focused on maintaining punctuality for trains.
+This indicates a strategy in handling rush periods by addressing operational challenges to prevent minor issues from causing major delays.
+Nevertheless the clustering of delays on weekdays, like Mondays and towards the weeks end implies that commuter patterns greatly impact the systems effectiveness.
+On Mondays, in particular there may be delays because of the influx of commuters heading back to work after the weekend break; while on Thursdays and Fridays we tend to see increased delays as folks get ready, for their weekend plans ahead of time.It might be helpful for the TTC to take note of these trends and adjust their resources accordingly to manage operations smoothly.
+
+The noticeable delays experienced on subway routes such as the Yonge–University and Bloor–Danforth lines demonstrate the difficulties in catering to populated areas with high passenger demand.These routes are lifelines for Torontos commuters and any disruptions here can cause far reaching effects throughout the city.Targeted approaches like upgrading signaling systems increasing train frequency and offering updates could be considered to improve service dependability.In addition, requiring traffic control to supervise site could also enhance the movement of passengers.
+
+In summary even though Toronto's subway system seems to handle peak hours despite the number of passengers.
+The examination reveals weaknesses during less busy times and, at specific stations and routes.
+To improve the situation the TTC could address these issues by improving services during off hours targeting stations with frequent delays and better managing resources on busy weekdays.
+These steps would greatly improve the system's effectiveness making the subway a more dependable and sturdy transportation option for the people of Toronto.
 

### 2024-09-27 scripts/00-simulate_data.R

@@ -11,16 +11,54 @@
 
 #### Workspace setup ####
 library(tidyverse)
+library(dplyr)
+library(ggplot2)
 
-#### Simulate data ####
-set.seed(123)
-n <- 100
+# Set random seed for reproducibility
+set.seed(096)
+
+# Define parameters
+lines <- c("Yonge-University", "Bloor-Danforth", "Sheppard", "Scarborough RT")
+stations <- c("Kennedy", "Kipling", "Eglinton", "Union", "Finch", "Bloor", "Spadina")
+days <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")
+
+# Define probabilities for delays on each line (based on real data insights)
+line_prob <- c("Yonge-University" = 0.4, "Bloor-Danforth" = 0.35, "Sheppard" = 0.15, "Scarborough RT" = 0.10)
+
+# Define a reasonable range for delay times (in minutes)
+min_delay <- 2   # Minimum delay
+max_delay <- 60  # Maximum delay
+
+# Number of simulated data points
+n_sim <- 2000  # Simulate 2000 subway delay events
+
+# Generate simulated data
 simulated_data <- data.frame(
-  date = seq(as.Date('2020-01-01'), by = "day", length.out = n),
-  transportation_usage = rpois(n, lambda = 200),
-  location = sample(c("Downtown", "Midtown", "Suburb"), n, replace = TRUE)
+  day = sample(days, n_sim, replace = TRUE),  # Randomly assign days of the week
+  line = sample(lines, n_sim, replace = TRUE, prob = line_prob),  # Randomly assign subway lines based on probabilities
+  station = sample(stations, n_sim, replace = TRUE),  # Randomly assign stations
+  delay_minutes = runif(n_sim, min = min_delay, max = max_delay)  # Randomly assign delay times between min and max delay
 )
 
-# Print sample data
-print(head(simulated_data))
+# Summarize the simulated data to inspect the first few rows
+head(simulated_data)
+
+# Visualize the total delay minutes per subway line
+ggplot(simulated_data, aes(x = line, y = delay_minutes, fill = line)) +
+  geom_boxplot() +
+  labs(title = "Simulated Subway Delay Data by Line", x = "Subway Line", y = "Delay Minutes") +
+  theme_minimal()
+
+# Visualize total delays by day of the week
[…]

### 2024-09-27 paper/paper.qmd
fourth update added sketches and updated discussion.

@@
 
-The noticeable delays experienced on subway routes such as the Yonge–University and Bloor–Danforth lines demonstrate the difficulties in catering to populated areas with high passenger demand.These routes are lifelines for Torontos commuters and any disruptions here can cause far reaching effects throughout the city.Targeted approaches like upgrading signaling systems increasing train frequency and offering updates could be considered to improve service dependability.In addition, requiring traffic control to supervise site could also enhance the movement of passengers.
-
 In summary even though Toronto's subway system seems to handle peak hours despite the number of passengers.

### 2024-09-27 paper/paper.qmd
Fifth update with font size and review of content

@@
 format: pdf
-fontsize: 12pt
 toc: true

### 2024-09-27 paper/paper.qmd
updated results paragraph

@@
 The overall analysis gives an overview of subway disruptions in Toronto for the year 2023.The frequency of delays is greatly impacted by the day of the week – particularly Mondays and the latter part of the workweek witness the highest delays.The Yonge-University and Bloor-Danforth lines face delays owing to their popularity and frequent train services.
-Specific stations like Kennedy and Eglinton serve as congestion points.
-Accumulate more delay minutes due to their vital positions in the transit system.
-During times when there are passengers traveling such as off hours or low traffic periods of the day there tends to be more delays in terms of total minutes being experienced due, to possibly having fewer trains running and maintenance work taking place.
-The thorough analysis offers an overview of subway disruptions in Toronto for the year 2023 with delays being significantly impacted by various factors and influences within the system.
-The analysis points out areas for operational enhancements to concentrate on such as improving off hours service and handling delays at busy stations like Kennedy and Eglinton to enhance the overall efficiency of the system better.
+Specific stations like Kennedy and Eglinton serve as transition points which accumulates more delay minutes due to their vital positions in the transit system.
+During times when there are passengers traveling such as off hours or low traffic periods of the day there tends to be more delays in terms of total minutes being experienced due to possibly having fewer trains running and maintenance work taking place.The analysis points out areas for operational enhancements to concentrate on such as improving off hours service and handling delays at busy stations like Kennedy and Eglinton to enhance the overall efficiency of the system better.
 Likewise tackling the issues on subway lines like YU and BD could lead to fewer disruptions especially during peak hours, with heavy passenger traffic.
